# Models for a single population

```{r}
#| warning: false
#| message: false
#| echo: false
library(tidyverse) # plotting, data organization
library(deSolve) # integrate differential equations
```

## A simple model for non-overlapping generations

Consider a population reproducing annually, with non-overlapping generations. Each individual produces $\rho$ individuals, which form the next generation. Then, we can write:

$$
x_{k+1} = \rho x_k
$$

where $x_k$ is the number of individuals/area at generation $k$. Naturally, we can iterate again and find the population density at time $x_{k+2}$:

$$
x_{k+2} = \rho x_{k+1} = \rho^2 x_{k}
$$ 

and in general:

$$
x_{n} = \rho^n x_0
$$

where $x_0$ is the initial population size, and $n$ is the number of generations since we started tracking the population. Naturally, we assume $x_0>0$, and thus the population grows whenever $\rho > 1$, shrinks whenever $\rho < 1$, and is constant whenever $\rho = 1$. Thus, armed with $x_0$ and $\rho$ we can project the population forward in time for an arbitrary number of generations. 

## Overlapping generations: exponential growth

Take the population at time $t$ and $t + \Delta t$; in the time interval $\Delta t$ the change in population size is $x(t + \Delta t) - x(t)$, and the average change per unit of time is $(x(t + \Delta t) - x(t))/\Delta t$. We can consider the limit in which the interval $\Delta t$ shrinks to zero:

$$
\lim_{\Delta t \to 0} \dfrac{x(t + \Delta t) - x(t)}{\Delta t} = \frac{dx(t)}{dt}
$$

In fact, this is exactly the definition of the derivative with respect to (w.r.t.) time of the function $x(t)$. A great contribution of Newton and Leibniz was to recognize that many natural phenomena can be described by simple differential equations---even though the equations describing the rate of change $dx(t)/dt$ are simple, the behavior of the solutions $x(t)$ can be incredibly complex.

In ecology and evolutionary biology, one encounters two main types of differential equations:

- Ordinary differential equations (ODEs) describe the rate of change of a quantity (in our case, population densities, $x(t)$) as a function of an independent variable (in our case, the time $t$), and can contain functions of the quantity, and its derivatives. If only the first derivative ($dx(t)/dt$) is included, they are called first-order ODEs. 

- Partial differential equations (PDEs) contain more than one independent variable (e.g., time and coordinates in space). Therefore, the quantity of interest is a multivariate function, such as $q(x,y,t)$, denoting the density of a population at the coordinates $x,y$ and time $t$. The partial derivatives with respect to the independent variables are considered, hence the name.

Many other types of differential equations (stochastic differential equations, SDEs; integro-differential equations, IDEs; etc.) are found in certain technical areas.

In this class, we will deal exclusively with ODEs of the form:

$$
\dfrac{dx(t)}{dt} = f(x(t))
$$

where the function $f(x(t))$ will model different ecological phenomena determining the growth and decline of populations. For example, take populations growing thanks to the per-capita birth rate $\beta$, and declining due to the death rate $\delta$:

$$
\dfrac{dx(t)}{dt} = \beta x(t) - \delta x(t) = (\beta - \delta) x(t) = \rho\, x(t)
$$

here the *intrinsic growth rate* $\rho$ represents the difference between birth and death rates. The parameter is taken to be independent of time, yielding an *autonomous* differential equation. An ODE is autonomous if it does not depend explicitly on the independent variable (in our case, time). I.e., if $dx/dt = f(x(t))$ the system is autonomous, while $dx/dt = g(x(t),t)$ is not.

::: {.callout-note collapse="true"}
## Separation of variables

A differential equation for $x(t)$ is called separable, if it can be written as:

$$
\frac{dx(t)}{dt} = g(t)\, h(x(t))
$$

where $g(t)$ is a function of $t$, and $h(x(t))$ a function of $x(t)$. As long as $h(x(t)) \neq 0$, we can formally write:

$$
\frac{1}{h(x(t))} dx(t) = g(t)\, dt
$$

We can now integrate both sides, obtaining 

$$
\begin{aligned}
\int \frac{1}{h(x(t))} dx(t) + \mathcal c_1 &= \int g(t)\, dt + \mathcal c_2\\
\int \frac{1}{h(x(t))} dx(t) &= \int g(t)\, dt + \mathcal c
\end{aligned}
$$

where $\mathcal c_1, \mathcal c_2$ and $\mathcal  c = \mathcal c_2 - \mathcal c_1$ are constants of integration, whose value can be set by considering the initial conditions. 

**Example** 

Consider:

$$
\frac{dx(t)}{dt} = \alpha
$$

with initial condition $x(0) = x_0$. Separate the variables and integrate:

$$
\begin{aligned}
d x(t) &= \alpha\, dt\\
\int d x(t) &= \alpha \int dt + \mathcal c\\
x(t) &= \alpha t + \mathcal c
\end{aligned}
$$

Now substitute the initial condition:

$$
x(0) = x_0 = \alpha 0 + c  = c
$$

Completing the solution:

$$
x(t) = x_0 + \alpha t
$$
:::

We can solve the differential equation:

$$
\begin{aligned}
\dfrac{dx}{dt} &= \rho\, x\\
\dfrac{1}{x}\,dx &= \rho\, dt
\end{aligned}
$$

Integrate both sides

$$
\begin{aligned}
\int \dfrac{1}{x}\,dx &= \int \rho\, dt\\
\log x &= \rho\, t + \mathcal c\\
x &= e^{\rho\, t  + \mathcal c}\\
x &= e^{\rho\, t} e^\mathcal c\\
\end{aligned}
$$

where $\mathcal c$ is a constant of integration. Then, we can plug in the initial condition: at $t=0$, the population is at density $x_0$:

$$
\begin{aligned}
x_0 &= e^{\rho\, 0}e^{\mathcal c}\\
x_0 &= e^\mathcal c
\end{aligned}
$$

We obtain the solution:

$$
x(t) = x(0) e^{\rho\, t}
$$

::: {.callout-tip collapse="true"}
## Integrating differential equations numerically

You can compute $x(t)$ for any differential equation (or system of differential equations) in `R` using numerical techniques. You need two parts:

First, we write a function defining the (system of) ODE(s). This function takes three arguments: `t`, the time, `x`, the state of the system, and `parms`, the parameters, and returns a `list` containing $dx/dt$. For the exponential growth, we can write:

```{r}
exponential_growth <- function(t, x, parms){
  dxdt <- x * parms$rho
  return(list(dxdt))
}
```

Because ecological models only make sense for positive population densities, we can set an arbitrarily small threshold, and consider the population extinct if it fall below the threshold:

```{r}
THRESH <- 10^-10

exponential_growth <- function(t, x, parms){
  if(x < THRESH) x <- 0
  dxdt <- x * parms$rho
  return(list(dxdt))
}
```

The second part of the code invokes the numerical integration function `ode`. For this function, we need to set a) initial conditions, b) determine at which times do we want to observe the state of the population, c) specify the name of the function to use, d) the parameters to use, and e) (optional) specify the method to use for the integration. The function `ode` is part of the package `deSolve`:

```{r}
x0 <- 1 # initial conditions
my_time <- seq(0, 5, by = 0.1) # we will observe the system at these times
parms <- list(rho = 1.05) # this is always a list of parameters
output <- ode(y = x0, # a) initial conditions
             times = my_time,  # b) time at which we observe the system
             func = exponential_growth, # c) function computing r.h.s. of ODEs
             parms = parms, # d) parameters of ODEs
             method = "ode45") # e) more on this later; this is a good general-purpose choice
```

The output is a matrix of class `deSolve`, containing the time in the first column, and the values of $x(t)$ in the second. For plotting, it is best to convert this into a data frame:

```{r}
output %>% 
  as.data.frame() %>% 
  ggplot() + aes(x = time, y = `1`) + 
  geom_point() + 
  geom_line() + 
  xlab("t") + 
  ylab(expression(x(t))) + 
  theme_bw()
```
:::


## Logistic equation

We have a population that grows exponentially when at very low densities, but experience a slow down in growth when densities are higher. The logistic growth equation is typically written as:

$$
\frac{d X}{d \tau} = X\left(\rho - \alpha X\right)
$$

where $\rho>0$ is the intrinsic growth rate of the population, and $\alpha>0$ that regulates the amount by which growth is decreased as $X$ increases. The equation has two equilibria: $X=0$ (absence of the population), and $\rho = \alpha X$, $X = \rho / \alpha = \kappa$, called the carrying capacity of the population.


::: {.callout-note collapse="true"}
## Chain rule

One of the most useful formulas from calculus. It relates the derivative of the composition of two differentiable functions with the derivatives of the functions. If $h(x) = g(u)$ and $u = f(x)$, then:

$$
\dfrac{dh}{dx} = \dfrac{dg}{du}\dfrac{du}{dx}
$$

As put by George Simmons [Calculus with Analytic Geometry (1985)]: *If a car travels twice as fast as a bicycle and the bicycle is four times as fast as a walking man, then the car travels 2 × 4 = 8 times as fast as the man.*

**Example**

Take 

$$
\dfrac{dx(t)}{dt} = x(t) f(x(t))
$$

write the ODE $d \log x(t) /dt$. We can write this as the derivative of the composition $\log x(t) = \log u$ and $u = x(t)$. Then, by chain rule:

$$
\begin{aligned}
\dfrac{d \log x(t)}{dt} &= \dfrac{d \log u}{du} \dfrac{du}{dt}\\
 &= \dfrac{1}{u} \dfrac{du}{dt}\\
 &= \dfrac{1}{x(t)} \dfrac{d x(t)}{dt}\\
 &= \dfrac{1}{x(t)}  x(t) f(x(t))\\
 &= f(x(t))
\end{aligned}
$$
showing that the per-capita rate of change:

$$
\dfrac{1}{x(t)} \dfrac{dx(t)}{dt} = \dfrac{\log x(t)}{dt}
$$
:::


::: {.callout-note collapse="true"}
## Nondimensionalization

Ecological models can have many parameters, and the goal of nondimensionalization is to rewrite the equations using as few parameters as possible, via a change of variables. Importantly, the new system and the original one are equivalent, and share the same dynamics. In fact, the trajectories of the original system can be reconstructed from those of the simpler system by inverting the transformation. 

To perform nondimensionalization:

- identify all the variables that depend on time ($X(\tau)$, $Y(\tau)$, etc.), as well as the independent variable (the time, $\tau$)
- replace each of them by a scaled version: $X = c_x x$, $Y = c_y y$, $t = c_t \tau$, in which $c_i$ are constants to be determined, and rewrite the equations for the new variables 
- choose the values for the constants to eliminate as many parameters as possible. Generally, one should be able to remove one parameter for each dependent variable, plus one parameter by scaling the time. 

## Examples

The exponential growth equation has a single parameter:

$$
\frac{dx}{d\tau} = \rho x
$$

one can define $t = c_1 \tau$, obtaining:

$$
\begin{aligned}
\frac{1}{c_1}\frac{dx}{dt} &= \rho x\\
\frac{dx}{dt} &= c_1 \rho x\\
\frac{dx}{dt} &= x
\end{aligned}
$$
where we have chosen $c_1 = 1/\rho$.
:::

We rewrite the equation for the logistic growth by choosing $x = c_1 X$ and $t = c_2 \tau$. Our equations become (use the chain rule):




::: {.callout-warning collapse="true"}
## Exercise 1

Take the equation for the logistic growth:

$$
\frac{dx}{dt} = x(1-x)
$$

- Solve the logistic growth model (the equation is separable). 
- Determine the location of all equilibria and their stability. 
- Write code to integrate the equation numerically.
- What happens if $x(0) > 1$, $x(0) = 1$, $0 < x(0) < 1$?
- (Bonus question) write a Lyapunov function for the logistic equation, and prove the global stability of the positive equilibrium.

:::
